---
title: "SEATRAC Introduction to R: Session 1"
author: "Shuyi Ma | Adapted from materials from Kim Dill-McFarland (https://bigslu.github.io/workshops/)"
date: "version `r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: no
  pdf_document:
    toc: yes
    toc_depth: '3'
editor_options:
  chunk_output_type: console
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#knitr::opts_chunk$set(fig.path = 'figures/')
##set working dir to project
#knitr::opts_knit$set(root.dir = #rprojroot::find_rstudio_root_file())
```

# Before starting
Please try to download and install R and RStudio prior to the workshop sessions with the following steps:

1. Download and install R: [download link](https://cran.r-project.org/)
2. Download and install Rstudio: [download link](https://posit.co/download/rstudio-desktop/)

You should also try to download the toy dataset that we will be using to demonstrate some concepts:
[Heart_Disease_100sampled.csv](https://github.com/Ma-Lab-Seattle-Childrens-CGIDR/SEATRAC_Rintro/raw/refs/heads/main/Heart_Disease_100sampled.csv)

# Overview
In this workshop, we'll introduce you to some R fundamentals:

* R data types (numbers, strings, logicals)
* Importing data from a spreadsheet into a data frame / tibble
* Subsetting data frames / tibbles
* Introduction to packages

# Basic data types
With R open, you will see a prompt '>'. Enter anything after this prompt, and R will try and evaluate it. R can evaluate 3 types of data:

* numbers and mathematical expressions
* quoted text (strings) 
* logical expressions (TRUE and FALSE are special values in R)

## Numbers
Calculator operations work with numbers.

```{r doubles}

# We are also assigning the answer to a variable we are calling a
a <- 1 + 1

# calling on a will give back the answer
print(a)

```

## Logicals

R calls conditionals "logicals". We can use operations that test whether something is TRUE or FALSE, which will output conditionals. 

```{r logicals}
3 < 5

4 < 2

#To test equality, you have to use two equals signs. You can use parentheses for operations
(3 + 3) == (2 + 4)

# To test for non-equality
4 != 3 
```

Note: when you want to compare 2 conditionals, you can test if both are true with the AND operator "&", and you can test if either one of the conditionals are true with the OR operator "|"
```{r AND/OR}
(3 < 5) & (4 < 2) #AND

(3 < 5) | (4 < 2) #OR
```

## Strings

R can work with text in the form of strings. Note that R will register anything put between "" or '' as a string, even numbers. But the operations R can perform on strings differ from those on numbers

```{r strings}
"Hello"

"Hello" == "Hello"

"Hello" == "Hello "

"Hello" == "hello"

```

```{r string error, echo=FALSE}
#'1'+'1'
#Error in "1" + "1" : non-numeric argument to binary operator
```

## Lists
Lists are a grouping of multiple entries of data. Can comprise of any data type
```{r list}
numlist = c(1,3,4,9)
numlist + 2
numlist > 3

# a shortcut
numrange = 1:4
print(numrange)

# a list of strings
wordlist = c('blue','red','white','green')
# can test if something is in a list:
'red' %in% wordlist

```

You can tell R to extract/subset specific elements of a list using square brackets 

```{r subsets}
numlist[2]
numlist[c(2,4)]
```

Importantly, you can subset based on conditionals, and embed conditional operations.
```{r conditional subsets}
print(numlist)
logiclist = c(TRUE,FALSE,TRUE,FALSE)
numlist[logiclist]
numlist[numlist < 5]
```


# Installing Packages in R
Many people have developed analysis tools for R that are not available in the base software. Instead, we can access these tools by downloading and installing packages. 

Installations only need to happen once.

```{r install}
#install.packages("tidyverse")
```

But you need to load the package every time you open R to access the tools provided in the package:

```{r library}
library(tidyverse)
```

# Data importing

We will use the tidyverse function `read_csv()` to import our spreadsheet into R as a tibble.

You will first need to navigate to the directory where you have downloaded our toy dataset, which can be done with the Dropdown menu: `Session > Set Working Directory`, or with the r command `setwd()`.
```{r import}
data = read_csv('Heart_Disease_100sampled.csv')
```

The spreadsheet dimensions can be queried by `dim`, `nrow`, and `ncol`
```{r data frame dimensions}
# tells number of rows (1st) and columns (2nd)
dim(data)

# nrow tells how many rows 
nrow(data)

#ncol tells how many columns
ncol(data)
```

R Studio: you can view a spreadsheet summary of the data with 
`View()`. `You can also see a summary of the data with `str()`
```{r viewing the data}
#View(data)

# viewing the top rows, columns 
str(data)
```

# Extracting data subsets
To select rows and columns together, use single brackets and two indices: [row, column].
```{r slicing data}
# Return the value at the first row, third column
data[1,4]

# You can also use lists:
# First 3 rows, first 2 columns
data[1:3, 1:2]

# all of columns in row 3
data[3,]

# all of rows in column 1
data[,1]
```

You can also call on specific columns by name.

```{r column slicing}
data$BMI
data[,'BMI']
```

Importantly, you can also subset based on conditionals

```{r conditional slicing}
HeartDiseaseRows = data$`Heart Disease Status` == 'Yes'
print(HeartDiseaseRows)
data[HeartDiseaseRows,'BMI']
```

## Example application: statistical testing with data subsets
This kind of subsetting enables you to perform comparative analyses.

As an example, we will perform a t-test, comparing the distribution of BMI in patients with and without heart disease. The R command for performing a standard Student's t-test is `t.test()`:

```{r ttest HD}
t.test(data[data$`Heart Disease Status` == 'Yes','BMI'],data[data$`Heart Disease Status` == 'No','BMI'])
```

# Getting help
To get more information on a function, use the ? operator. This brings up a help page describing how the function works. These pages can be pretty dense and technical, even for longtime R users, so don't be discouraged if it's hard to understand
```{r help ex}
?t.test()
```


To see an example of how a function is used, try the 'example' function. Not every function has examples though!
```{r example}
example(t.test)
```
Note: Searching with a specific question will also return many useful tutorials online.

### Bonus example: a Linear Model associating BMI with different factors

Clinical data are often analyzed by linear regression analyses. In R, the command to perform an ordinary least squares linear model is `lm()`. The `summary()` function reports several result outputs from the linear model. 

```{r linear model BMI}
IndepVar = data$BMI
FactorGender = as.factor(data$Gender)
FactorSmoking = as.factor(data$Smoking)
FactorExercise = as.factor(data$`Exercise Habits`)

BMImodel = lm(IndepVar ~ FactorGender + FactorSmoking + FactorExercise)

summary(BMImodel)
```

